Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-05-03 paper.tex
Updates from Overleaf

A further challenge is that OCF is infeasible if the number of required codewords is too large, again because OCF requires constructing a matrix whose rows and columns are indexed by the codewords and observed sequences.
Given that storing and manipulating matrices greater than $10^5 \times 10^5$ is currently not feasible, a setting where $\operatorname{I}[\past, Z] > \log 10^5 \approx 11.5$ cannot be captured with OCF.

@@

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|\finitepast),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $\finitepast$ into the parameters of the distribution $P_\phi(\cdot|\finitepast)$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} \\
	&= \frac{\E_{\finitepast} P(\finitefuture, Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} \\
	&=^{(*)} \frac{\E_{\finitepast}   P(\finitefuture|\finitepast)     P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition $Z \bot \finitefuture | \finitepast$.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[\finitepast, Z]$ of the code $Z \sim P_\phi(Z|\finitepast)$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:prd-loss}) in a way amenable to using function approximators:

\begin{corollary}[Main Result]
The predictive rate-distortion objective~(\ref{eq:prd-loss})
\begin{equation}\tag{\ref{eq:prd-loss}}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \operatorname{I}[ \finitepast, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} [ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.

From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-loss}) by setting $Z$ to be the code generated by the encoder $\phi$.
%The second proposition guarantees that, for a solution to  (\ref{eq:prd-loss}), 
The rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, \finitepast] =   \mathbb{E}_{\finitepast}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
and the prediction loss is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[\finitefuture|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. %H[\finitefuture|\finitepast] % \operatorname{H}[\finitefuture]
\end{equation}
\end{corollary}
\begin{proof}
TODO write
\end{proof}

@@

We first consider the problem of predicting English on the level of Part-of-Speech (POS) tags, using the Universal POS tagset \citep{petrov-universal-2012}. 
This is a simplified setting where the vocabulary is small (20 word types), and one can hope that OCF will produce reasonable results.
We use the English portions of the Universal Dependencies Project~\citep{nivre-universal-2017} tagged with Universal POS Tags \citep{petrov-universal-2012}, consisting of approximately 586K words.
We used the training portions to estimate NPRD and OCF, and the validation portions to estimate rate-distortion.
We used NPRD to generate 350 codebooks for values of $\lambda$ sampled from [0,0.4].
We were only able to run OCF for $M \leq 3$, as the number of sequences exceeds $10^4$ already at $M=4$.

>>>>

A further challenge is that OCF is infeasible if the number of required codewords is too large, again because it requires constructing a matrix whose rows and columns are indexed by the codewords and observed sequences.
Given that storing and manipulating matrices greater than $10^5 \times 10^5$ is currently not feasible, a setting where $\operatorname{I}[\past, Z] > \log 10^5 \approx 11.5$ cannot be captured with OCF.

@@

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|\finitepast),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $\finitepast$ into the parameters of the distribution $P_\phi(\cdot|\finitepast)$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} = \frac{\E_{\finitepast} P(\finitefuture, Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} =^{(*)} \frac{\E_{\finitepast}   P(\finitefuture|\finitepast)     P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition $Z \bot \finitefuture | \finitepast$.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[\finitepast, Z]$ of the code $Z \sim P_\phi(Z|\finitepast)$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:prd-loss}) in a way amenable to using function approximators, which is our main theoretical result, and the foundation of our proposed estimation method:

\begin{corollary}[Main Result]
The predictive rate-distortion objective~(\ref{eq:prd-loss})
\begin{equation}\tag{\ref{eq:prd-loss}}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \operatorname{I}[ \finitepast, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q} \left[\mathbb{E}_{\finitepast, \finitefuture}	\left[-	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} [ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.

From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-loss}) by setting $Z \sim P_\phi(\cdot|\finitepast)$.
%The second proposition guarantees that, for a solution to  (\ref{eq:prd-loss}), 
The rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, \finitepast] =   \mathbb{E}_{\finitepast}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
and the prediction loss is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[\finitefuture|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. %H[\finitefuture|\finitepast] % \operatorname{H}[\finitefuture]
\end{equation}
\end{corollary}

\begin{proof}
By the two propositions, the term inside the minimization in (\ref{eq:bound}) is an upper bound on (\ref{eq:prd-loss}), and takes on that value if and only if $P_\phi(\cdot|\finitepast)$ equals the distribution of the $Z$ optimizing (\ref{eq:prd-loss}), and $\psi, q$ are as described in those propositions.
\end{proof}

@@

We first consider the problem of predicting English on the level of Part-of-Speech (POS) tags, using the Universal POS tagset \citep{petrov-universal-2012}. 
This is a simplified setting where the vocabulary is small (20 word types), and one can hope that OCF will produce reasonable results.
We use the English portions of the Universal Dependencies Project~\citep{nivre-universal-2017} tagged with Universal POS Tags \citep{petrov-universal-2012}, consisting of approximately 586K words.
We used the training portions to estimate NPRD and OCF, and the validation portions to estimate the rate-distortion curve.
We used NPRD to generate 350 codebooks for values of $\lambda$ sampled from [0,0.4].
We were only able to run OCF for $M \leq 3$, as the number of sequences exceeds $10^4$ already at $M=4$.
